Analysis of celiac disease

### 2018-08-21 5-results-celiac.Rmd
celiac results with grid search

```{r}
library(tidyverse)
myggplot <- function(..., coeff = 1) {
  ggplot(...) + bigstatsr::theme_bigstatsr(size.rel = coeff)
} 
```

@@

```{r, cache=FALSE}
# Read file, see "Preprocessing" notebook
library(bigsnpr)
celiac <- snp_attach("backingfiles/celiacQC.rds")
G <- celiac$genotypes
CHR <- celiac$map$chromosome
POS <- celiac$map$physical.pos
NCORES <- nb_cores()
y <- celiac$fam$affection - 1

@@

```{r, cache=TRUE}
# Compute PCA
obj.svd <- snp_autoSVD(G, CHR, POS, ncores = NCORES)
```

@@

```{r}
pAUC <- function(pred, target, p = 0.1) {
  val.min <- min(target)
  q <- quantile(pred[target == val.min], probs = 1 - p)
  ind <- (target != val.min) | (pred > q)
  tryCatch(bigstatsr::AUC(pred[ind], target[ind]) * p, 
           error = function(e) p^2 / 2)  # all same prediction
}
plot_density_AUC <- function(pred, true = y[ind.test]) {
  true_fct <- forcats::fct_recode(as.factor(true), "Control" = "0", "Case" = "1")
  data.frame(Score = pred, Status = true_fct) %>%
  myggplot() +
  geom_density(aes(Score, fill = Status), alpha = 0.4) +
  ggtitle(sprintf("AUC = %s%% | pAUC = %s", 
                  round(100 * AUC(pred, true), 2),
                  round(pAUC(pred, true), 4))) + 
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
}
(p1 <- plot_density_AUC(prs[, which.max(aucs)]))
```

@@

```{r, cache=TRUE}
# Train logit-simple
system.time(
  logit <- big_spLogReg(X = G, y01.train = y[ind.train], 
                        ind.train = ind.train, 
                        covar.train = obj.svd$u[ind.train, ],
                        ncores = NCORES, alpha = 0.5)
)
# Get K = 10 predictions/individuals
preds1 <- predict(logit, X = G, ind.row = ind.test, 
                  covar.row = obj.svd$u[ind.test, ])
(nb.pred1 <- sum(rowSums(sapply(logit, function(l) l$beta.X != 0)) > 0))
(aucs1 <- apply(preds1, 2, AUC, target = y[ind.test]))
# Combining all K predictions
preds1.comb <- rowMeans(preds1)
AUC(preds1.comb, y[ind.test])
```

```{r}
(p2 <- plot_density_AUC(preds1.comb))
```

@@

```{r, cache=TRUE}
system.time(
  logit2 <- big_spLogReg(X = G2, y01.train = y[ind.train], 
                         ind.train = ind.train, 
                         covar.train = obj.svd$u[ind.train, ],
                         ncores = NCORES, alphas = 0.5)
)

preds2 <- predict(logit2, X = G2, ind.row = ind.test, 
                  covar.row = obj.svd$u[ind.test, ])
(nb.pred2 <- sum(rowSums(sapply(logit2, function(l) l$beta.X != 0)) > 0))
(aucs2 <- apply(preds2, 2, AUC, target = y[ind.test]))
# Combining all K predictions
preds2.comb <- rowMeans(preds2)
AUC(preds2.comb, y[ind.test])
```

```{r}
(p3 <- plot_density_AUC(preds2.comb))
```

@@

```{r, fig.asp=1, fig.width=8}
library(plotROC)
scores_tidy <- tibble(
  d = y[ind.test], 
  "PRS-max"      = prs[, which.max(aucs)],
  "logit-simple" = preds1.comb,
  "logit-triple" = preds2.comb
) %>%
  gather(key = "Method", value = "Score", -d)

@@

bigstatsr:::MY_THEME(
  ggplot(scores_tidy, aes(d = d, m = Score, color = Method, linetype = Method)) +
    style_roc(xlab = "1 - Specificity", ylab = "Sensitivity")
) +
  geom_roc(n.cuts = 0, size = 2) +
  theme(legend.position = c(0.7, 0.3), legend.key.width = unit(4, "line")) +
  coord_equal() +
  scale_color_manual(values = methods.color)
```

@@

## Top scores

First, we get the most significant SNP.

```{r}
# Most significant SNP
theone <- which.max(-predict(gwas.train.gc))
ind0 <- which(G[ind.test, theone] == 0)
(M <- length(ind0))
table(Geno = G[, theone], forcats::fct_recode(as.factor(y), Control = "0", Case = "1"))
```

```{r}
# one geno
mean(y[ind.test[ind0]]) 
# prs max
indPRS <- order(prs[, which.max(aucs)], decreasing = TRUE)[1:M]
mean(y[ind.test[indPRS]])
# logit-simple (CMSA)
indLS <- order(preds1.comb, decreasing = TRUE)[1:M]
mean(y[ind.test[indLS]])
# logit-triple (CMSA)
indLT <- order(preds2.comb, decreasing = TRUE)[1:M]
mean(y[ind.test[indLT]])
```

@@

#### PRS

```{r}
PRS <- function(G, infos.chr, infos.pos, 
                pheno.all, covar.all, 
                ind.train, ind.test) {

  timing <- system.time({

    # GWAS
    gwas.train <- big_univLogReg(
      G, pheno.all[ind.train], ind.train = ind.train, 
      covar.train = covar.all[ind.train, , drop = FALSE], 
      ncores = nb_cores()
    )
    gwas.train.gc <- snp_gc(gwas.train)

    # Clumping on the test set
    ind.keep <- snp_clumping(G, infos.chr = infos.chr,
                             ind.row = ind.test,
                             thr.r2 = 0.2, 
                             S = abs(gwas.train.gc$score),
                             size = 500,
                             is.size.in.bp = TRUE,
                             infos.pos = infos.pos,
                             ncores = nb_cores())

    # PRS
    thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
    lpS <- -predict(gwas.train.gc)
    nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
    prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
                   ind.test = ind.test,
                   ind.keep = ind.keep,
                   lpS.keep = lpS[ind.keep], 
                   thr.list = thrs)
  })[3]

  ind.best <- which.max(apply(prs, 2, AUC, target = pheno.all[ind.test]))

  tibble(
    method = "PRS-max",
    eval = list(cbind(prs[, ind.best], pheno.all[ind.test])),
    timing = timing,
    nb.preds = nb.pred[ind.best]
  )
}
```

#### Logistic regression

```{r}
logit.CMSA <- function(G, pheno.all, covar.all, ind.train, ind.test, method, print_all = FALSE) {

  timing <- system.time({

    cmsa.logit <- big_spLogReg(X = G, y01.train = pheno.all[ind.train], 
                               ind.train = ind.train, 
                               covar.train = covar.all[ind.train, , drop = FALSE],
                               ncores = NCORES, alphas = 0.5)

    preds.all <- predict(cmsa.logit, X = G, ind.row = ind.test, 
                         covar.row = covar.all[ind.test, , drop = FALSE])
    if (print_all) print(preds.all)
    preds <- rowMeans(preds.all)
  })[3]

  tibble(
    method   = method, 
    eval     = list(cbind(preds, pheno.all[ind.test])),
    timing   = timing,
    nb.preds = sum(rowSums(sapply(cmsa.logit, function(x) x$beta.X) != 0) > 0)
  )
}
```

@@

  # Get results from all methods
  res <- bind_rows(
    PRS(G, CHR, POS, pheno.all, covar.all, ind.train, ind.test),
    logit.CMSA(G,  pheno.all, covar.all, ind.train, ind.test, "logit-simple"),
    logit.CMSA(G2, pheno.all, covar.all, ind.train, ind.test, "logit-triple")
  )
  saveRDS(res, file = res.file)
}
```

```{r}
pAUC <- function(pred, target, p = 0.1) {
  val.min <- min(target)
  q <- quantile(pred[target == val.min], probs = 1 - p)
  ind <- (target != val.min) | (pred > q)
  tryCatch(bigstatsr::AUC(pred[ind], target[ind]) * p, 
           error = function(e) p^2 / 2)  # all same prediction
}

format_results <- function(results) {
  results %>%
    as_tibble() %>%
    mutate(
      AUC = map_dbl(eval, ~ bigstatsr::AUC(.x[, 1], .x[, 2])),
      pAUC = map_dbl(eval, ~ pAUC(.x[, 1], .x[, 2]))
    )
}

boot <- function(x, n = 1e5, f = mean) {
  sd(replicate(n, f(sample(x, replace = TRUE))))
}

list.files(path = "results6/", full.names = TRUE) %>%
  map_df(readRDS) %>%
  format_results() %>%
  group_by(method) %>%
  summarize_at(c("AUC", "pAUC", "nb.preds", "timing"), function(x) {
    glue::glue("{signif(mean(x), 3)} ({signif(boot(x), 3)})")
  }) %>%
  print() %>%
  xtable::xtable()
```

>>>>

```{r}
library(tidyverse)
library(bigsnpr)
library(pkg.paper.PRS)
```

@@

```{r, cache=FALSE}
# Read file, see "Preprocessing" notebook
celiac <- snp_attach("backingfiles/celiacQC.rds")
G <- celiac$genotypes
CHR <- celiac$map$chromosome
POS <- celiac$map$physical.pos
NCORES <- nb_cores()
y <- celiac$fam$affection - 1
obj.svd <- readRDS("backingfiles/PCA.rds")

@@

@@

```{r}
pAUC <- pkg.paper.PRS:::pAUC

plot_density_AUC <- function(pred, true = y[ind.test]) {
  true_fct <- forcats::fct_recode(as.factor(true), "Control" = "0", "Case" = "1")
  data.frame(Score = pred, Status = true_fct) %>%
  ggplot() +
  geom_density(aes(Score, fill = Status), alpha = 0.4) +
  ggtitle(sprintf("AUC = %s%% | pAUC = %s", 
                  round(100 * AUC(pred, true), 2),
                  round(pAUC(pred, true), 4))) + 
    theme_bigstatsr() +
    theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
}
(p1 <- plot_density_AUC(prs[, which.max(aucs)]))
```

@@

```{r, cache=TRUE}
# Train logit-simple
system.time(
  logit <- big_spLogReg(X = G, y01.train = y[ind.train], 
                        ind.train = ind.train, 
                        covar.train = obj.svd$u[ind.train, ],
                        ncores = NCORES, alphas = 1)
)
# Get K = 10 predictions/individuals
preds1 <- predict(logit, X = G, ind.row = ind.test, 
                  covar.row = obj.svd$u[ind.test, ])
(nb.pred1 <- sum(rowSums(sapply(logit, function(l) l$beta.X != 0)) > 0))
(aucs1 <- AUC(preds1, target = y[ind.test]))
```

```{r}
(p2 <- plot_density_AUC(preds1))
```

@@

```{r, cache=TRUE}
system.time(
  logit2 <- big_spLogReg(X = G2, y01.train = y[ind.train], 
                         ind.train = ind.train, 
                         covar.train = obj.svd$u[ind.train, ],
                         ncores = NCORES, alphas = 1)
)

preds2 <- predict(logit2, X = G2, ind.row = ind.test, 
                  covar.row = obj.svd$u[ind.test, ])
(nb.pred2 <- sum(rowSums(sapply(logit2, function(l) l$beta.X != 0)) > 0))
(aucs2 <- AUC(preds2, target = y[ind.test]))
```

```{r}
(p3 <- plot_density_AUC(preds2))
```

@@

```{r, fig.asp=1, fig.width=8}
library(plotROC)
scores_tidy <- tibble(
  d = y[ind.test], 
  "PRS-max"      = prs[, which.max(aucs)],
  "logit-simple" = preds1,
  "logit-triple" = preds2
) %>%
  gather(key = "Method", value = "Score", -d)

@@

ggplot(scores_tidy, aes(d = d, m = Score, color = Method, linetype = Method)) +
  style_roc(xlab = "1 - Specificity", ylab = "Sensitivity") +
  theme_bigstatsr() +
  geom_roc(n.cuts = 0, size = 2) +
  theme(legend.position = c(0.7, 0.3), legend.key.width = unit(4, "line")) +
  coord_equal() +
  scale_color_manual(values = methods.color)
```

@@

@@

@@

  # Get results from all methods
  res <- bind_rows(
    filter(PRS(G, CHR, POS, pheno.all, covar.all, ind.train, ind.test), thr.r2 == 0.2),
    logit.CMSA(G,  pheno.all, covar.all, ind.train, ind.test, "logit-simple"),
    logit.CMSA(G2, pheno.all, covar.all, ind.train, ind.test, "logit-triple")
  )
  res$pheno <- list(pheno.all[ind.test])

  saveRDS(res, file = res.file)
}
```

```{r}
list.files(path = "results6/", full.names = TRUE) %>%
  map_df(readRDS) %>%
  as_tibble() %>%
  mutate(
    AUC = map2_dbl(pred, pheno, ~bigstatsr::AUC(.x, .y)),
    pAUC = map2_dbl(pred, pheno, ~pAUC(.x, .y)),
    nb.preds = map_int(set, length)
  ) %>%
  group_by(method) %>%
  summarize_at(c("AUC", "pAUC", "nb.preds", "timing"), function(x) {
    glue::glue("{signif(mean(x), 3)} ({signif(boot(x), 3)})")
  }) %>%
  print() %>%
  xtable::xtable()
```
